---
title: "SCEPTRE analysis of sample Gasperini data"
author: "Gene Katsevich"
date: "November 16, 2023"
output: 
  bookdown::pdf_document2:
    toc: false
urlcolor: blue
---

```{r, echo = FALSE}
# set chunk options to cache
knitr::opts_chunk$set(cache = TRUE, fig.pos = "H")
```

In this document, we will apply `sceptre` to analyze the sample Gasperini data.

# Load libraries and data

```{r, message = FALSE}
# load libraries
library(MuData)
library(dplyr)
library(ggplot2)
library(conflicted)
library(kableExtra)
library(sceptre)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::filter)
conflicts_prefer(base::unname)
```

```{r}
# read in the MuData object
gasperini_data_dir <- "~/code/research/Pipeline_Gasperini_2019/"
gasperini_data_fp <- paste0(
  gasperini_data_dir, 
  "/mudata/Gasperini_2019_sample_pilot.h5mu"
)
gasperini_data <- readH5MU(gasperini_data_fp)
```

# Import data into SCEPTRE

The first step is to get the data into a form that `sceptre` expects; see [Import data from a collection of R objects](https://timothy-barry.github.io/sceptre-book/import-data.html#import-data-from-a-collection-of-r-objects) in the manual.

```{r}
scRNA_data <- gasperini_data@ExperimentList$scRNA
guides_data <- gasperini_data@ExperimentList$guides
response_matrix <- scRNA_data@assays@data@listData[[1]]
grna_matrix <- guides_data@assays@data@listData[[1]]
extra_covariates <- scRNA_data@colData |>
  as.data.frame() |> 
  select(percent_mito) # batch_number omitted because it is constant in these data
moi <- "high"
response_names <- scRNA_data@rowRanges@elementMetadata$feature_name |> 
  as.character()
grna_target_data_frame <- guides_data@rowRanges@elementMetadata |>
  as.data.frame() |>
  rename(grna_id = feature_name, grna_target = target_elements, chr = guide_chr, 
         start = guide_start, end = guide_end) |>
  select(grna_id, grna_target, chr, start, end)
```

```{r, echo = FALSE}
shorten_grna_name <- function(grna_name){
 strsplit(x = grna_name, split = "_sgrna_")[[1]][1]
}
rownames(grna_matrix) <- sapply(rownames(grna_matrix), shorten_grna_name) |> unname()
grna_target_data_frame <- grna_target_data_frame |>
  mutate(grna_id = sapply(as.character(grna_id), shorten_grna_name))
```

Next, we import this data into a `sceptre_object`.

```{r}
sceptre_object <- import_data(
  response_matrix = response_matrix,
  grna_matrix = grna_matrix,
  grna_target_data_frame = grna_target_data_frame,
  moi = moi,
  extra_covariates = extra_covariates,
  response_names = response_names
)
```

We can print this `sceptre_object` to get basic information about the data:

```{r}
sceptre_object
```

# Set analysis parameters

The next step is to set the analysis parameters.

```{r}
# construct positive control pairs
response_ids <- rownames(response_matrix)
on_target_response_idx <- which(response_names %in% grna_target_data_frame$grna_target)
positive_control_pairs <- data.frame(
  grna_target = response_names[on_target_response_idx], 
  response_id = response_ids[on_target_response_idx]
)
# no discovery pairs, since no enhancer-targeting perturbations present
discovery_pairs <- data.frame(grna_target = character(), response_id = character())
# set analysis parameters
sceptre_object <- sceptre_object |>
  set_analysis_parameters(
    discovery_pairs = discovery_pairs,
    positive_control_pairs = positive_control_pairs,
    side = "left"
  )
```

# Assign gRNAs to cells

Next, we assign gRNAs to cells. Prior to this, it is useful to visualize the gRNA count distributions. Here, we do so for three gRNAs:

```{r, fig.width = 8, fig.height = 2.5, out.width = "\\linewidth", fig.align='center'}
# plot gRNA count distributions
grnas_to_plot <- c("NEK2|1", "ODC1|1", "SEPT11|1")
sceptre_object |>
  plot_grna_count_distributions(grnas_to_plot = grnas_to_plot)
```

The gRNA distributions are nicely bimodal, with 5 being a threshold to divide gRNAs with background reads from those actually present in a cell (this was Gasperini's original choice). Within `sceptre`, we can either use the thresholding method or a more sophisticated mixture method to assign gRNAs to cells. The mixture method adaptively sets a different threshold for different gRNAs in different cells.

```{r, results = "hide"}
# assign gRNAs to cells via thresholding
sceptre_object <- sceptre_object |> 
  assign_grnas(method = "thresholding", threshold = 5)

# alternatively, assign gRNAs to cells via mixture
sceptre_object_mixture <- sceptre_object |> 
  assign_grnas(method = "mixture")
```

We can visualize the results of these two gRNA assignments (Figures \@ref(fig:thresholding) and \@ref(fig:mixture)).

```{r thresholding, out.width = "0.75\\linewidth", fig.align='center', fig.cap = "Thresholding-based gRNA assignment"}
plot(sceptre_object, grnas_to_plot = grnas_to_plot)
```

```{r mixture, out.width = "0.75\\linewidth", fig.align='center', fig.cap = "Mixture-based gRNA assignment"}
plot(sceptre_object_mixture, grnas_to_plot = grnas_to_plot)
```

On these data, the two methods appear to give similar results.

# Quality control

The next step is quality control. We can visualize covariates like the number of genes expressed in a cell ("Response N nonzero") and the total number of UMIs in a cell ("Response N UMIs").

```{r, out.width = "0.75\\linewidth", fig.align='center'}
sceptre_object |> plot_covariates()
```

We can remove cells with extreme values of these covariates via `run_qc`; we choose the default values of the QC thresholds, depicted in the figure above.

```{r}
sceptre_object <- sceptre_object |> run_qc()
```

# Positive control analysis

Finally, we analyze the positive control pairs via `run_power_check()`, of which there are only 8 in this sample dataset.

```{r, results = "hide", warning = FALSE}
sceptre_object <- sceptre_object |> run_power_check()
```

The next figure shows the positive control p-values. 

```{r, fig.width = 3, fig.height = 3, fig.align='center'}
sceptre_object |> 
  plot() +
  labs(title = "Positive control p-values")
```

Finally, we can print a table of the positive control p-values:

```{r}
sceptre_object |>
  get_result(analysis = "run_power_check") |>
  select(grna_target, response_id, p_value, log_2_fold_change) |>
  mutate(p_value = format(p_value, scientific = TRUE, digits = 2)) |>
  kable(format = "latex",
        row.names = NA,
        booktabs = TRUE,
        digits = 2,
        caption = "Positive control p-values",
        linesep = '') |>
  kable_styling(latex_options = "hold_position")
```

The positive control p-values are quite small, which is a good sign.