---
title: "SCEPTRE interoperability with MuData"
author: "Gene Katsevich"
date: "December 2, 2023"
output: 
  bookdown::pdf_document2:
    toc: false
urlcolor: blue
---

```{r, echo = FALSE}
# set chunk options to cache
knitr::opts_chunk$set(cache = TRUE, fig.pos = "H")
```

In this document, we present the `sceptreIGVF` package ([GitHub](https://github.com/Katsevich-Lab/sceptreIGVF)), which provides an interface between the `sceptre` package and `MuData` objects. At present, the `sceptreIGVF` package provides the following two functions:

- `convert_mudata_to_sceptre_object()` converts a `MuData` object to a `sceptre_object`
- `assign_grnas_sceptre()` inputs a `MuData` object, uses `sceptre` to assign gRNAs, and returns a new `MuData` object with another assay containing the gRNA assignments

# Converting `MuData` to `sceptre_object`

```{r, message = FALSE}
# load libraries
library(sceptreIGVF)
library(MuData)
library(MultiAssayExperiment)
```

Let us take the Gasperini pilot data as an example:
```{r}
# read in the MuData object
gasperini_data_dir <- "~/code/research/Pipeline_Gasperini_2019/"
gasperini_data_fp <- paste0(
  gasperini_data_dir, 
  "/mudata/Gasperini_2019_sample_pilot.h5mu"
)
gasperini_mudata <- readH5MU(gasperini_data_fp)
gasperini_mudata
```

We can convert the `MuData` object to a `sceptre_object` using `convert_mudata_to_sceptre_object()`:
```{r, message = FALSE}
gasperini_sceptre_object <- convert_mudata_to_sceptre_object(gasperini_mudata)
gasperini_sceptre_object
```
We could then apply the `sceptre` pipeline to this `sceptre_object`. However, we might be interested in integrating individual `sceptre` modules into our Nextflow pipeline. For this reason, we would like to have functions wrapping around `sceptre` that operate directly on `MuData` objects.

# gRNA assignment based on `MuData` objects

Going back to the initial `MuData` object, let us use `assign_grnas_sceptre()` to assign gRNAs based on the thresholding method:
```{r}
gasperini_mudata_updated <- assign_grnas_sceptre(
  mudata = gasperini_mudata, 
  method = "thresholding", 
  threshold = 5
)
```

The result is another `MuData` object with an additional assay containing the gRNA assignments:
```{r}
gasperini_mudata_updated
```
In the context of a Nextflow pipeline, the resulting `MuData` object could be written to disk using `MuData::writeH5MU()` and then read in by the next process in the pipeline.

The function `sceptreIGVF::assign_grnas_sceptre()` has the same gRNA assignment options as `sceptre::assign_grnas()`: thresholding, maximum, and mixture. For documentation of these options, see the corresponding portion of the `sceptre` vignette [here](https://katsevich-lab.github.io/sceptre/articles/sceptre.html#sec-sceptre_assign_grnas). 
